### 2025-02-18 paper/sections/other/constraints.qmd
vamos

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments. Below we derive the analytical results in @thm-mtblty.

@@

This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$. In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.

By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$. In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

>>>>

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments. Below we derive the analytical results in @prp-mtblty.

@@

This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will decrease $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$. In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.

By the same rationale, the contrastive divergence penalty can generally be expected to increase $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$. In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

### 2025-02-18 paper/sections/related.qmd
vamos

This interpretation of the link between explainability through counterfactuals on one side, and robustness to adversarial examples on the other, is backed by empirical evidence. @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization. @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function. In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *POLYJUICE*, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through *POLYJUICE* counterfactuals improves robustness in a number of NLP tasks. @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models. Specifically, they propose to proceed as follows: firstly, they identify training samples that are subject to high predictive uncertainty; secondly, they generate counterfactual explanations for those samples; and, finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.

>>>>

This interpretation of the link between explainability through counterfactuals on one side, and robustness to adversarial examples on the other, is backed by empirical evidence. @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization. @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function.

In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *Polyjuice*, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks. @balashankar2023improving also use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models. Specifically, they propose to proceed as follows: firstly, they identify training samples that are subject to high predictive uncertainty; secondly, they generate counterfactual explanations for those samples; and, finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.
